Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-11-19 Paper/ms.tex
Eccentricity prior

\subsection{Parameters and Priors}
To fit a planet model to RV data, we need parameters to describe
the properties of each planet. For simplicity, we describe each planet by
five parameters: $T_i$, the orbital period, $A_i$, the amplitude (in metres
per second) of the RV signal, $\phi_i$, the phase of the signal, $v_i$, an
``initial velocity'' parameter describing the eccentricity of the orbit
(see Section~\ref{sec:orbits}), and $\theta_i$, a ``viewing angle'' parameter
describing the position of the observer with respect to the major axis of the
star's elliptical trajectory.

@@

The unknown parameters are:
\begin{eqnarray}
\left\{N, \boldsymbol{\alpha}, \{\boldsymbol{\psi}\}_{i=1}^N, y_0, s, \nu\right\}
\end{eqnarray}
where $N$ is the number of planets,
$\boldsymbol{\alpha} = \{\mu_T, \sigma_T, \mu_A\}$ are the
hyperparameters used to define the prior for the properties of the planets,
and $\psi_i = \{T_i, A_i, \phi_i, v_i, \theta_i\}$
are the properties of planet $i$.
The parameter $y_0$ describes a DC offset in the data, and $s$ and $\nu$ are
parameters allowing for a non-gaussian distribution for the noise in the
measurements. When the model-predicted RV signal is $m(t)$, the sampling
distribution for each measurement was a student-$t$ distribution with
scale parameter $\sqrt{\sigma_i^2 + s_0^2}$ and shape parameter $\nu$. In the limit
$(s_0 \to 0, \nu \to \inf)$ the sampling distribution reduces to the standard
assumption of a normal distribution with standard deviation $\sigma_i$ known
from the error bars in the data set.

@@

\begin{table*}
\begin{tabular}{|l|l|l|}
\hline
Quantity	&	Meaning		& Prior\\
\hline
{\bf Hyperparameters}	&	\\
$N$		& Number of planets	& Uniform$(\{0, 1, ..., N_{\rm max}\})$\\
$\mu_T$		&	Median orbital period	& $\log(\mu_T) \sim$ Cauchy$(5.9, 1)T(-15.3, 27.1)$\\
$w_T$		&	Diversity of orbital periods & $w_T \sim$ Uniform$(0.1, 3)$\\
$\mu_A$		&	Mean amplitude	& $\log(\mu_A) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
\hline
{\bf Planet Parameters}\\
$T_i$		&	Orbital period	&	$\log(T_i) \sim $ Biexponential($\log(\mu_T), w_T$)\\
$A_i$		&	Amplitude of signal	& Exponential$(\textnormal{mean}=\mu_A)$\\
$\phi_i$	&	Phase of signal	&	Uniform($0, 2\pi$)\\
$v_i$		&	Initial speed of orbit	&	Uniform$(0.4, 1)$\\
$\theta_i$	&	Viewing angle	&	Uniform$(0, 2\pi)$\\
\hline
{\bf Other}\\
$m_0$		&	Constant DC offset	&	Uniform$(m_{0, \rm min}, m_{0, \rm max})$\\
$\sigma_{\rm extra}$	& ``Extra noise'' parameter	& $\log(\sigma) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
$\nu$		& Shape parameter for $t$-distribution for noise & $\log(\nu) \sim$ Uniform$(\log(0.01), \log(1000))$\\
\hline
{\bf Data}\\
$Y_i$		& Radial velocity measurements	&
		Student-$t\left(m(t_i), \sqrt{\sigma_i^2 + \sigma_{\rm extra}^2}, \nu\right)$
\end{tabular}
\caption{All of the prior distributions in our Bayesian model.
The priors for the planet parameters are defined conditional on the values
of the hyperparameters. Uniform priors were used for parameters like phases.
For parameters where a rough initial guess is possible, heavy-tailed Cauchy
distributions were used so this information could be taken into account
in a non-dogmatic way. Time units are in days and amplitude units are
in metres per second. The maximum number of planets, $N_{\rm max}$, was set
to 10. The prior limits for $m_0$ were set to the minimum and maximum value
in the dataset, which is not strictly a valid strategy.\label{tab:priors}}
\end{table*}

@@

Consider a test particle moving in the $x$-$y$ plane under the influence of a
point mass at the origin. The motion of the test particle represents the
reflex motion of the host star orbiting around the center of mass of the
system. The equations of motion for the particle are:
\begin{eqnarray}
\frac{d^2x}{dt^2} &=& -\frac{x}{r^3} \\
\frac{d^2y}{dt^2} &=& -\frac{y}{r^3} \\
\end{eqnarray}
where $r = \sqrt{x^2 + y^2}$. The solutions to this system of equations are elliptical orbits
with the focus at the origin. We set the initial position to $(1, 0)$, and
the initial velocity to $(0, v)$ where $v \in [0.4, 1]$.
If $v=1$, the orbit is circular and as $v$ decreases the orbit becomes more
elliptical. For trial values of $v$ ranging from 0.4 to 1 in steps of 0.005,
we calculated the orbit, and saved the
velocities $\frac{dx}{dt}$ and $\frac{dy}{dt}$ as a function of time to disk.
These saved orbits were used as a lookup table for constructing the expected
signal $y(t)$ due to a single planet.
Because of the initial conditions, the simulated orbits were all horizontally
aligned. If the observer is located on the $x$-axis a large distance
from the origin, they will measure $m(t) = \dot{x}(t)$. However, if the
observer is located at an angle $\theta$ with respect to the $x$-axis, then
the radial velocity measured will instead be
$m(t) = \cos(\theta)\dot{x}(t) + \sin(\theta)\dot{y}(t)$.
Since the our orientation with respect to the orbits is unknown, each planet
requires a ``viewing angle'' parameter $\theta$.
Although we parameterised the orbits using the initial velocity $v$,
the standard eccentricity $\epsilon$ can be recovered using the relation
$\epsilon = 1 - v^2$.

By precomputing a set of orbits before running the MCMC, we are able to
do $\sim$ 15,000 likelihood evaluations per second per CPU core.

>>>>

\subsection{Parameters and Priors}
To fit a planet model to RV data, we need parameters to describe
the properties of each planet. For simplicity, we describe each planet by
five parameters: $T_i$, the orbital period, $A_i$, the amplitude (in metres
per second) of the RV signal, $\phi_i$, the phase of the signal, $e_i$, the eccentricity of the orbit, and $\theta_i$, a ``viewing angle'' parameter
describing the position of the observer with respect to the major axis of the
star's elliptical trajectory.

@@

The unknown parameters are:
\begin{eqnarray}
\left\{N, \boldsymbol{\alpha}, \{\boldsymbol{\psi}\}_{i=1}^N, y_0, s, \nu\right\}
\end{eqnarray}
where $N$ is the number of planets,
$\boldsymbol{\alpha} = \{\mu_T, \sigma_T, \mu_A\}$ are the
hyperparameters used to define the prior for the properties of the planets,
and $\psi_i = \{T_i, A_i, \phi_i, e_i, \theta_i\}$
are the properties of planet $i$.
The parameter $y_0$ describes a DC offset in the data, and $s$ and $\nu$ are
parameters allowing for a non-gaussian distribution for the noise in the
measurements. When the model-predicted RV signal is $m(t)$, the sampling
distribution for each measurement was a student-$t$ distribution with
scale parameter $\sqrt{\sigma_i^2 + s_0^2}$ and shape parameter $\nu$. In the limit
$(s_0 \to 0, \nu \to \inf)$ the sampling distribution reduces to the standard
assumption of a normal distribution with standard deviation $\sigma_i$ known
from the error bars in the data set.

@@

\begin{table*}
\begin{tabular}{|l|l|l|}
\hline
Quantity	&	Meaning		& Prior\\
\hline
{\bf Hyperparameters}	&	\\
$N$		& Number of planets	& Uniform$(\{0, 1, ..., N_{\rm max}\})$\\
$\mu_T$		&	Median orbital period	& $\log(\mu_T) \sim$ Cauchy$(5.9, 1)T(-15.3, 27.1)$\\
$w_T$		&	Diversity of orbital periods & $w_T \sim$ Uniform$(0.1, 3)$\\
$\mu_A$		&	Mean amplitude	& $\log(\mu_A) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
\hline
{\bf Planet Parameters}\\
$T_i$		&	Orbital period	&	$\log(T_i) \sim $ Biexponential($\log(\mu_T), w_T$)\\
$A_i$		&	Amplitude of signal	& Exponential$(\textnormal{mean}=\mu_A)$\\
$\phi_i$	&	Phase of signal	&	Uniform($0, 2\pi$)\\
$e_i$		&	Eccentricity of orbit	&	Beta$(1, 3.1)T(0, 0.8)$\\
$\theta_i$	&	Viewing angle	&	Uniform$(0, 2\pi)$\\
\hline
{\bf Other}\\
$m_0$		&	Constant DC offset	&	Uniform$(m_{0, \rm min}, m_{0, \rm max})$\\
$\sigma_{\rm extra}$	& ``Extra noise'' parameter	& $\log(\sigma) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
$\nu$		& Shape parameter for $t$-distribution for noise & $\log(\nu) \sim$ Uniform$(\log(0.01), \log(1000))$\\
\hline
{\bf Data}\\
$Y_i$		& Radial velocity measurements	&
		Student-$t\left(m(t_i), \sqrt{\sigma_i^2 + \sigma_{\rm extra}^2}, \nu\right)$
\end{tabular}
\caption{All of the prior distributions in our Bayesian model.
The priors for the planet parameters are defined conditional on the values
of the hyperparameters. Uniform priors were used for parameters like phases.
For parameters where a rough initial guess is possible, heavy-tailed Cauchy
distributions were used so this information could be taken into account
in a non-dogmatic way. Time units are in days and amplitude units are
in metres per second. The maximum number of planets, $N_{\rm max}$, was set
to 10. The prior limits for $m_0$ were set to the minimum and maximum value
in the dataset, which is not strictly a valid strategy.\label{tab:priors}}
\end{table*}

@@

Consider a test particle moving in the $x$-$y$ plane under the influence of a
point mass at the origin. The motion of the test particle represents the
reflex motion of the host star orbiting around the center of mass of the
system. The equations of motion for the particle are:
\begin{eqnarray}
\frac{d^2x}{dt^2} &=& -\frac{x}{r^3} \\
\frac{d^2y}{dt^2} &=& -\frac{y}{r^3} \\
\end{eqnarray}
where $r = \sqrt{x^2 + y^2}$. The solutions to this system of equations are elliptical orbits
with the focus at the origin. We set the initial position to $(1, 0)$, and
the initial velocity to $(0, v)$ where $v \in [0.4, 1]$.
If $v=1$, the orbit is circular and as $v$ decreases the orbit becomes more
elliptical. For trial values of $v$ ranging from 0.4 to 1 in steps of 0.005,
we calculated the orbit, and saved the
velocities $\frac{dx}{dt}$ and $\frac{dy}{dt}$ as a function of time to disk.
These saved orbits were used as a lookup table for constructing the expected
signal $y(t)$ due to a single planet.
Because of the initial conditions, the simulated orbits were all horizontally
aligned. If the observer is located on the $x$-axis a large distance
from the origin, they will measure $m(t) = \dot{x}(t)$. However, if the
observer is located at an angle $\theta$ with respect to the $x$-axis, then
the radial velocity measured will instead be
$m(t) = \cos(\theta)\dot{x}(t) + \sin(\theta)\dot{y}(t)$.
Since the our orientation with respect to the orbits is unknown, each planet
requires a ``viewing angle'' parameter $\theta$.
The eccentricity of the orbit, in terms of $v$, is $e = 1 - v^2$.
By precomputing a set of orbits before running the MCMC, we are able to
do $\sim$ 15,000 likelihood evaluations per second per CPU core.
